---
title: "EDA of RNA-Seq pseudocounts from Jaffe _et al._"
author: "[Nima Hejazi](http://nimahejazi.org)"
date: '`r format(Sys.time(), "%Y %b %d (%a), %H:%M:%S")`'
output: pdf_document
theme: journal
highlight: haddock
---

```{r preliminaries, message=FALSE, warning=FALSE, echo=FALSE}
rm(list = ls())
proj_dir <- paste0(path.expand("~"), "/git_repos/neurodevstat")
source(paste0(proj_dir, "/munge/01_getCounts.R"))
source(paste0(proj_dir, "/munge/02_makeDesign.R"))
source(paste0(proj_dir, "/munge/03_voomLimma.R"))

library(reshape2)
library(ggbiplot)
library(wesanderson)
pal1 <- wes_palette("Chevalier", n = 2, type = "discrete")
pal2 <- wes_palette("Darjeeling", type = "continuous")
```

```{r FUN_multiplot, message=FALSE, warning=FALSE, echo=FALSE}
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

We provide several plots produced as part of the exploratory data analysis (EDA)
procedures performed in examining this data set, and comment on the information
provided by the EDA procedure.

Note that the plots are generated from a matrix of "pseudocounts," created
from the use of the pseudoalignment algorithm proposed in [Bray _et al._
(2016)](http://www.nature.com/nbt/journal/v34/n5/full/nbt.3519.html). Since the
alignment procedure differs slightly from that used in standard alignment
software (_e.g._, bowtie), the counts are not integers. For any concerns about
the validity of this alignment procedure, please consult the Bray _et al._ paper
or the [website of the Kallisto
pseudoaligner](https://pachterlab.github.io/kallisto/about).

```{r boxplots_rawCounts, message=FALSE, warning=FALSE, echo=FALSE}
counts.raw <- reshape2::melt(pseudocounts_filtered)
p1 <- ggplot(data = counts.raw, aes(x = variable, y = value)) +
  geom_boxplot() + xlab("") + ylab("Pseudocount") +
  ggtitle("Boxplot: raw pseudocount") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r boxplots_normalized, message=FALSE, warning=FALSE, echo=FALSE}
pseudocounts_filtNorm <- normalizeBetweenArrays(pseudocounts_filtered,
                                                method = "scale")
counts.norm <- reshape2::melt(data.frame(pseudocounts_filtNorm))
p2 <- ggplot(data = counts.norm, aes(x = variable, y = value)) +
  geom_boxplot() + xlab("") + ylab("Pseudocount") +
  ggtitle(paste("Boxplot: normalized pseudocount",
                "\n (normalization method: median)")) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r print_boxplots, message=FALSE, warning=FALSE, echo=FALSE}
multiplot(p1, p2, cols = 2)
```

The boxplot of the raw (pseudo)counts (graph on the left) indicates considerable
variability across the twelve samples, with __SRR1554536__ containing a number
of (pseudo)counts that could be considered outliers. Among the other eleven
samples, there is a fair amount of variability, though this degree of variation
is generally expected and acceptable. Based on the empirical distributions of
the raw (pseudo)counts, some form of normalization is warranted. To remove a
significant degree of unwanted variation, we perform __median normalization__,
a procedure which forces the medians of the distributions across samples to be
the same. Careful examination of the normalized (pseudo)counts (graph on the
right) suggests that the normalization procedure appears to force the empirical
distributions to be similar, indicating that the procedure removes a deal of
technical variability, while preserving variation that might be considered to be
due to biological variability between samples. In spite of the normalization
procedure, the empirical distribution of sample __SRR1554536__ still appears to
be problematic, suggesting that this sample may suffer from issues of quality
and should likely be downweighted in later steps involving modeling.

```{r pca, message=FALSE, warning=FALSE, echo=FALSE}
pseudocounts_pc <- prcomp(t(pseudocounts_filtNorm),
                          center = TRUE, scale. = TRUE)
```

```{r pca_plot_1, message=FALSE, warning=FALSE, echo=FALSE}
g1 <- ggbiplot(pseudocounts_pc, obs.scale = 1, var.scale = 1, 
               groups = ifelse(design_full$type == 1, "Fetal", "Adult"),
               var.axes = FALSE, ellipse = FALSE, circle = FALSE)
g1 <- g1 + scale_color_manual(values = pal1, name = "type")
g1 <- g1 + ggtitle("PCA Biplot of Samples")
```

```{r pca_plot_2, message=FALSE, warning=FALSE, echo=FALSE}
g2 <- ggbiplot(pseudocounts_pc, obs.scale = 1, var.scale = 1, 
               groups = ifelse(design_full$sex == 1, "Female", "Male"),
               var.axes = FALSE, ellipse = FALSE, circle = FALSE)
g2 <- g2 + scale_color_manual(values = pal1, name = "sex")
g2 <- g2 + ggtitle("PCA Biplot of Samples")
```

```{r pca_plot_3, message=FALSE, warning=FALSE, echo=FALSE}
g3 <- ggbiplot(pseudocounts_pc, obs.scale = 1, var.scale = 1, 
               groups = ifelse(design_full$race == 1, "HISP", "AA"),
               var.axes = FALSE, ellipse = FALSE, circle = FALSE)
g3 <- g3 + scale_color_manual(values = pal1, name = "race")
g3 <- g3 + ggtitle("PCA Biplot of Samples")
```

```{r pca_plot_4, message=FALSE, warning=FALSE, echo=FALSE}
g4 <- ggbiplot(pseudocounts_pc, obs.scale = 1, var.scale = 1,
               groups = cut(design_full$RIN,
                            seq(min(design_full$RIN) - 0.5,
                                max(design_full$RIN) + 0.5, 1.25)),
               var.axes = FALSE, ellipse = FALSE, circle = FALSE)
g4 <- g4 + scale_color_manual(values = pal2, name = "RIN")
g4 <- g4 + ggtitle("PCA Biplot of Samples")
```

```{r print_pca_plots, message=FALSE, warning=FALSE, echo=FALSE}
multiplot(g1, g2, g3, g4, cols = 2)
```

The four principal component biplots presented above show a clear separation
between the samples after projection into the space spanned by the first two
principal components. A cursory examination of the biplot in the top left corner
of the plot matrix indicates a clear separation between the samples based on the
phenotype of interest (__fetal__ vs. __adult__); furthermore, it is clear from
the biplot that the samples are clearly separated in even the subspace of the
first principal component (x-axis). The remaning principal component biplots
indicate that the samples are fairly balanced across phenotype measures that are
not directly of interest but may contribute to downstream analytic results. In
particular, the biplot in the top right corner of the plot matrix indicates that
nearly all of the samples originated from individuals of the same race, while
the biplot in the bottom left corner of the plot matrix shows that each grouping
of samples (across the main phenotype of interest) is balanced across sexes.
The biplot in the bottom right corner of the plot matrix indicates that the
considerable variation in quality of the samples (as measured by __RIN__) is
fairly balanced across the two groups. On a final note, it is worth bringing
attention to the fact that a single sample is separated from the rest after
projection into the subspace of the first two principal components, and that it
is this sample that corresponds to __SRR1554536__, which was noted to contain a
number of outlying counts in the boxplots displayed in the previous section.